Comparing summed probability distributions of shoreline and radiocarbon dates from the Mesolithic Skagerrak coast of Norway

### 2023-06-11 analysis/paper/paper.Rmd
Monte Carlo simulation, shoreline dates

---
title: "Comparing summed probability distributions of shoreline and radiocarbon dates from the Stone Age on the Norwegian Skagerrak coast"
# author: 
# - 'Isak Roalkvam'
# - 'University of Oslo, Institute of Archaeology, Conservation and History'
  # - IsakRoalkvam:
  #     email: isak.roalkvam@iakh.uio.no
  #     institute: [iakh]
      # correspondence: true
#   - name: Steinar Solheim
#     email: steinar.solheim@khm.uio.no
#     institute: [mch]
#     correspondence: false
institute:
  - iakh: Institute of Archaeology, Conservation and History, University of Oslo
#   - mch: Museum of Cultural History, University of Oslo
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::pdf_document2:
      fig_caption: yes
      toc: false
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      # - --lua-filter=../templates/scholarly-metadata.lua
      # - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
# abstract: |
#   Text of abstract
# keywords: |
#   keyword 1; keyword 2; keyword 3
# highlights: |
#   These are the highlights. 
---

@@

```{r packages, echo = FALSE}
library(here)
library(dplyr)
library(kableExtra)
library(sf)
library(shoredate)
```

@@

# Introduction
Population size is regarded as one of the primary drivers of cultural variation, and is of critical importance to our understanding of past human societies [@shennan2000]. The frequency distribution of radiocarbon dates has been used extensively as one proxy for past relative population sizes [e.g. @crema2022; @french2021], including in Norwegian Stone Age archaeology [@solheim2018a; @solheim2020a; @nielsen2019; @nielsen2021; @bergsvik2021; @jorgensen2020; @lundstrom2021, see also @persson2009]. The potentially immense value of insights into past population dynamics, combined with the ubiquity of radiocarbon dates and the relative ease with which these can now be processed within what has been termed a dates as data methodology has undoubtedly contributed to the popularity of the approach. Several limitations and forms of criticism have, however, been directed at these procedures. Some of the objections are of a methodological nature, while others pertain to the underlying logic and the degree to which there is likely to be a direct connection between the frequency of ^14^C-dates and population dynamics [@carleton2021; @torfing2015]. What appears to be agreed upon by practitioners and critics alike is that radiocarbon dates are best analysed in this manner when compared and contrasted to other proxies for past population dynamics and other variables that might impact these [@french2016; @palmisano2017; @palmisano2021; @bergsvik2021]. This paper reports on and begins to unpack the relationship between two measures that have been linked to relative population size in the context of the Mesolithic Skagerrak coast in south-eastern Norway, namely summed probability distribution of calibrated radiocarbon dates and the summed probability distribution of shoreline dated sites (hereafter RSPD and SSPD, respectively).

@@

To what degree the radiocarbon record is determined by past population numbers might vary both geographically and chronologically, based on variation in investigatory and taphonomic factors [@surovell2009b; @bluhm2019], as well as cultural processes within prehistoric populations. One example of the latter is the difference that might exist between farmer and forager populations, where @freeman2018 have suggested that an increased per capita energy consumption introduced with farming means that ^14^C-dates should not be weighted equally when making relative population estimates across such populations. Similarly, while site counts have also been invoked for the analysis of past population dynamics, these are likely to be impacted by factors such as land-use and mobility patterns, and settlement nucleation and dispersion [@palmisano2017]. However, these could be considered theoretical issues that have implications for how fluctuations in these proxies should be interpreted and possibly weighted differently according some criteria. Before any such fluctuations are given any substantive interpretation, however, there are a host of methodological issues that have to be considered.

The most critical of these follow from the fact that the summation of the probabilities associated with the dates for the SPDs is not a statistically coherent procedure. This is because the summed probabilities can no longer been seen as probabilities, but rather represents the combination of events and uncertainties, making the two indistinguishable, and rendering the interpretation of the resulting sum difficult [@blackwell2003; @crema2022]. As @timpson2021[2] put it: 'the SPD is not *the* single best explanation of the data, nor even *a* explanation of the data, but rather a conflation of many possible explanations simultaneously, each of which is mired by the artefacts inherited from the calibration wiggles.' The SPD is not a model. It is the combined representation of a range of possible explanations for the data -- the frequency of dated events combined with the variable uncertainty associated with these [@carleton2021]. This means that an SPD cannot be directly analysed to draw inferences on population dynamics, nor can it be directly compared to other time-series data [@carleton2018; @timpson2021]. While this problem can never be entirely resolved, a range of approaches have been developed in an attempt to work around this issue.

The most commonly applied of these is a null-hypothesis significance testing approach by means of Monte Carlo simulation, as introduced by @shennan2013 and later expanded upon by @timpson2014. This works by comparing the observed RSPD with a series of simulated RSPDs, generated from a null-model. These null-models are typically a uniform, exponential or logistic distribution. These are chosen *a priori* with reference to common long-term population dynamics and are parametrised by fitting the models by means of regression or maximum likelihood estimation (MLE). The result from these simulations are then used to create a 95% critical envelope representing the null model. The proportion of the observed RSPD that falls outside this envelope is used to estimate a global p-value indicating whether the null model can be rejected. In the case that it can, the portions of the observed RSPD that falls outside this envelope can subsequently be interpreted as representing potentially meaningful demographic events, relative to the null model. However, care has to be taken in how these are interpreted. First, this follows from the fact that 5% of the deviations from the critical envelope can be expected to be random, and there is no way to know which deviations this pertains to. Secondly, a for example exponential null model fit to the data is only one of an infinite set of exponential models with different growth rates that could be used. While a model fit by means of MLE will likely have a reasonable growth rate, and by extension exclude many other exponential fits as likely to explain the data, this can be difficult to determine. Finally, the p-value only indicates whether or not the null-model as a whole can be rejected as an explanation of the data, and does not provide statistical justification for interpreting local deviations themselves as meaningful demographic signals, as has often been the case [see review by @crema2022; @timpson2021].

@@

The procedure outlined above represent the standard approach when modelling population dynamics in archaeology. This can be helpful for determining whether the population behaviour has followed the trajectories of standard theoretical models, and is a useful first step here as few assumptions concerning the behaviour and overall development of number of sites over time could be made. However, standard models frequently used to evaluate RSPDs are consistently being rejected in the literature, and deviations from these rejected null-models do not in themselves allow for any interpretations of the deviations, nor a subsequent direct analysis of the SPDs themselves. As a result, and by improving on @goldstein, @timpson2021 have recently suggested an alternative framework for modelling past population dynamics based RSPDs, published with the accompanying R package  *ADMUR*. This involves fitting continuous piece-wise linear (CPL) models in addition to the standard models to the data, subjecting these to a procedure of model comparison, and finally testing the best model using the Monte Carlo approach outlined above. The benefit of the CPL models is that these define discrete periods between hinge points that mark points in time in which the trajectory of the population proxy changes direction, thus allowing for the identification of key events with a conservative description of population dynamics between these. The approach thus also provides a clearer inferential justification for a further analysis and substantive interpretation of the identified best model. Following the evaluation of standard models as outlined above, the data were therefore subsequently subjected to the approach presented by @timpson2021.

@@

Sites surveyed by means of test-pitting between Horten and Arendal were initially retrieved from the national heritage database Askeladden [@directorate2018], totalling at `r nrow(surveyed)` records. The records were then manually reviewed and given a quality score based on the criteria in Table \@ref(tab:tab1), indicating the degree to which the spatial location and extent of the sites is believed to be represented in the geometries available in the database [see also @roalkvam2020]. All sites with a quality score of 4 or worse were excluded from further analysis. Any sites situated at elevations that result in a shoreline date earlier than 9469 BCE were then excluded. This marks the latest start-date among the employed displacement curves, and no sites are yet verifiably older than around 9300 BCE [@glorstad2016]. Data on excavated sites was originally compiled for @roalkvam2023 and has been compared with site data as listed in @damlien2021 and @nielsen2022. Only excavated sites with available spatial data in Askeladden or local databases at the Museum of Cultural History of the University of Oslo were included in the analysis. The 102 excavated sites in the dataset without relevant ^14^C-dates and that were originally shoreline dated in the reports were included in the SSPD along with the retained surveyed sites. This gave a total of  shoreline dated sites in the final SSPD.

The borders of the municipalities within which the shoreline dated sites are located were used to limit the radiocarbon sample. Radiocarbon dates were taken from @roalkvam2023 and @solheim. Dates done on food crusts were then excluded due the issue of marine reservoir effects [@nielsen2019, 83], resulting in a final total of ^14^C-dates. Following both from the point made by @freeman concerning the comparison of radiocarbon dates from populations with different , as noted above,  and from the fact that shoreline dating appears less reliable after 4000 BCE and looses its utility c. 2500 BCE, the Mesolithic was the main focus of this analysis. Consequently, calibrated ^14^C-dates with a probability falling later than 2500 BCE were excluded from the analysis. Dates falling after 4000 BCE and the first small-scale introduction of agriculture in south-eastern Norway were retained to account for edge-effects, but these results should be treated with care for the reasons noted above.

All analyses done in this study were performed using the R programming language [@rcoreteam]. Underlying data and programming code used for the paper is available in a version-controlled online repository at X. This is structured as a research compendium following @marwick2018b, to allow for reproducibility of the results [@Marwick2017]. Analysis of the ^14^C-dates were done using the R package *rcarbon* [@crema2021]. The R package *shoredate* was used for performing and handling the shoreline dating of sites. Finally, the package *ADMUR* [@timpson2021] was used for model comparison and selection, as well as for estimating .

@@

# Map prepared in 01_shoreline_spd.R 
knitr::include_graphics(here("analysis/figures/map.png"))
```

@@

The RSPD was first subjected to the standard null-hypothesis testing approach through Monte Carlo simulation, as introduced above [see @shennan2013; @timpson2014], by fitting an exponential, logistic and uniform model to the observed RSPD. Following @timpson2021, the model parameters were identified by MLE search, using the differential evolution optimization algorithm *DEoptimR*. The Monte Carlo simulations were then performed using the *modelTest()* function from *rcarbon*. For each model, a series of individual calendar years are drawn from the distribution, with replacement, the number of which equals the number of bins in the observed RSPD. These are then 'uncalibrated' to a single value on the ^14^C scale and a random error from among the observed errors is added to the date. These are then calibrated back to the calendar scale and finally summed. Here, this procedure was repeated 10000 times for each null model. The 2.5th and 97.5th quantile of the resulting summed probability for each year across all simulations are then retrieved to create the 95% critical envelope with which to compare the observed RSPD. The degree to which the observed RSPD deviates from the critical envelope is then used to calculate a global p-value, indicating whether or not the null can be rejected.

@@

Summing the probability of the shoreline dated sites and the model-fitting procedures followed the same structure as that for radiocarbon dates and was partly based on re-purposed programming code from *rcarbon*. However, idiosyncrasies in the dating method did necessitate some adjustments. To illustrate this, the procedure for shoreline dating a single site, as suggested in @roalkvam2023, is provided in Figure \@ref(fig:shoredate). Four geologic reconstructions of shoreline displacement in the region lays the foundation for the method as implemented here. These shoreline displacement curves are from Horten [@romundset2021], Porsgrunn [@sørensen2014; @sørensen2014b; @sorensen2023], Tvedestrand [@romundset2018; @romundset2018b] and Arendal [@romundset2018b], each associated with a shoreline isobase along which the trajectory for relative sea-level change has been the same [see @svendsen1987]. The first step in the dating procedure is to interpolate the shoreline displacement to the location to be dated. This is done by inverse distance weighting [e.g. @conolly2020], interpolating the sea-level to the location of a site to be dated based on it distance from the isobases of the displacement curves. The analysis in @roalkvam2023 found that site phases ^14^C-dated to before 2500 BCE were found to have a likely elevation above sea-level that was reasonably approximated by a gamma distribution with shape =  and scale = . When shoreline dating, the gamma distribution is here sequentially stepped through at increments of 0.1m and transferred to the calendar year scale by uniformly distributing the probability across the years in the range between the lower and upper limit of the interpolated displacement curve. This procedure gives the the shoreline date of a site [see @roalkvam2023 for details]. Given that the shoreline displacement curves have no inversions and should therefore before commutative [cf. @weninger2015, 545], each shoreline date is normalised to sum to unity. To reduce the computational cost of the simulation procedures, the gamma distribution is here stepped through at increments of 0.1 m and the calendar scale is kept at a resolution of 5 years.

```{r shoredate, echo = FALSE, fig.cap = "A) Example location relative to the isobases of the displacement curves.  B) The geologic displacement curves and the curve interpolated to the example location. C) Resulting shoreline date in light grey on the x-axis. The black line underneath marks the 95\\% HDR. The dashed line marks the elevation of the example location. The exponential function on the y-axis decays with ratio $\\lambda$ and represents the likely elevation of a site above sea-level when it was in use. ", out.width = "500px", fig.align="center", regfloat = TRUE}
target_site <- sites[1,]
target_plot(target_site)

```

<!-- The SSPD was subjected to the same procedure for CPL model fitting and comparison as teh RSPD above. Dates with 50% probability distribution  -->

In Figure  the shoreline dated sites are plotted according to their elevation. However, due to the variable uplift rates within the study region, the same elevation does not necessarily equate to the same date, and so this does not directly tell us much about their temporal distribution. As the trajectory for sea-level regression is starker towards the north-east, the same elevation at a location to the south-west implies a younger date that one towards the north-east. If the elevation of the sites is instead transformed to point-estimated shoreline dates by finding the mean shoreline date for the sites, without accounting for the likely distance to the shoreline, the temporal distribution instead looks like the one provided in Figure. This is similar to how shoreline dates have been treated when the frequency distribution of shoreline dated sites has been investigated in the past [@breivik; @solheim; @fossum; @mjærum; @tallavaara]. However, apart from not accounting for the distance between the site and the sea, this does not properly account for the uncertainty associated with the dates, and is thus similar to summarising point-estimates of ^14^C-dates. This problem is further accentuated by the fact that shoreline dating, as implemented here, is associated with more uncertainty than ^14^C-dates. Additionally, given that the displacement curves have a more rapid drop early in the Holocene and at locations further to the north-east in the study area, this means that this also results in both a geographical and temporal bias in the final results. The SSPD resulting from using the probabilistic method of shoreline dating is given in Figure, which clearly demonstrates that accounting for these dimensions gives a different result compared to the traditional approach.

@@

```{r incpolys, echo = FALSE, fig.cap = "A) Density of included surveyed sites and excavated sites dated by means of shoreline dating (n = 921) as distributed across the 2km wide line segments that run perpendicular to the shoreline gradient. B) The displacement curves interpolated to the centre of each segment for use in the Monte Carlo simulations below.", out.width = "500px", fig.align="center", regfloat = TRUE}
knitr::include_graphics(here("analysis/figures/incpolys.png"))
```

Following from the commutative nature of the shoreline dates, the Monte Carlo simulation is based on drawing a sample of calendar dates from the observed date range in the SSPD, equalling the number of shoreline dated sites, where the probability of drawing any individual year is determined by the null model of choice. This is equivalent to the sampling method *calsample* from *rcarbon* [@crema2021]. Each sampled date is then assigned one of the pre-interpolated displacement curves, the probability of which is weighted by the density of observed sites within each 2km interval. The calendar date is then 'uncalibrated' -- to follow the RSPD terminology -- to an elevation range from which a single elevation value is then drawn with uniform probability, using intervals of 5 cm. As shoreline dating is here done with a gamma distribution with the same parameters across all sites, there is no equivalent of the error term for ^14^C-dates that determine the shape of the Gaussian distribution associated with the ^14^C age. Consequently, the elevation value retrieved by 'uncalibrating' the shoreline date was shoreline dated using the displacement curve for the relevant 2 km interval with the same gamma distribution for all samples. Having dated the number of samples equalling the number of shoreline dated sites, these were then summed, and the entire process repeated a total of 10000 times. The 97.5% highest and 2.5% lowest summed probability at each interval of 5 years across all simulations was then retrieved to create the 95% critical envelope, and the global p-value found in the same manner as for the RSPDs. This was done using re-purposed code from *rcarbon*.

>>>>

---
title: "Comparing summed probability distributions of shoreline and radiocarbon dates from the Mesolithic Skagerrak coast of Norway"
# author: 
# - 'Isak Roalkvam'
# - 'University of Oslo, Institute of Archaeology, Conservation and History'
  # - IsakRoalkvam:
  #     email: isak.roalkvam@iakh.uio.no
  #     institute: [iakh]
      # correspondence: true
#   - name: Steinar Solheim
#     email: steinar.solheim@khm.uio.no
#     institute: [mch]
#     correspondence: false
institute:
  - iakh: Institute of Archaeology, Conservation and History, University of Oslo
#   - mch: Museum of Cultural History, University of Oslo
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::pdf_document2:
      fig_caption: yes
      toc: false
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      # - --lua-filter=../templates/scholarly-metadata.lua
      # - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
header-includes:
  \usepackage{placeins}
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
# abstract: |
#   Text of abstract
# keywords: |
#   keyword 1; keyword 2; keyword 3
# highlights: |
#   These are the highlights. 
---

@@

```{r packages, echo = FALSE}
library(here)
library(dplyr)
library(ggplot2)
library(patchwork)
library(kableExtra)
library(sf)
library(shoredate)
```

@@

# Introduction
Population size is regarded as an important driver of cultural variation, and is of critical importance to our understanding of past human societies [e.g. @french2021; @shennan2000]. The frequency distribution of radiocarbon dates has been used extensively as one proxy for past relative population sizes [e.g. @crema2022; @french2021], including in Norwegian Stone Age archaeology [@solheim2018; @solheim2020a; @nielsen2019; @nielsen2021; @bergsvik2021; @jorgensen2020; @lundstrom2021; see also @persson2009]. The potentially immense value of insights into past population dynamics, combined with the ubiquity of radiocarbon dates and the relative ease with which these can now be processed within what has been termed a dates as data methodology has undoubtedly contributed to the popularity of the approach. Several limitations and forms of criticism have, however, been directed at these procedures. Some of the objections are of a methodological nature, while others pertain to the underlying logic and the degree to which there is likely to be a direct connection between the frequency of ^14^C-dates and population dynamics [@carleton2021; @torfing2015]. What appears to be agreed upon by practitioners and critics alike is that radiocarbon dates are best analysed in this manner when compared and contrasted to other proxies for past population dynamics and other variables that might impact these [@french2016; @palmisano2017; @palmisano2021; @bergsvik2021]. This paper reports on and begins to unpack the relationship between two measures that have been linked to relative population size in the context of the Mesolithic Skagerrak coast in south-eastern Norway, namely summed probability distribution of calibrated radiocarbon dates and the summed probability distribution of shoreline dated sites (hereafter RSPD and SSPD, respectively).

@@

To what degree the radiocarbon record is determined by past population numbers might vary both geographically and chronologically, based on variation in investigatory and taphonomic factors [@surovell2009b; @bluhm2019], as well as cultural processes within prehistoric populations. One example of the latter is the difference that might exist between farmer and forager populations, where @freeman2018 have suggested that an increased per capita energy consumption introduced with farming means that ^14^C-dates should not be weighted equally when making relative population estimates across such populations. Similarly, while site counts have also been invoked for the analysis of past population dynamics, these are likely to be impacted by factors such as land-use and mobility patterns, and settlement nucleation and dispersion [@palmisano2017]. However, these could be considered theoretical issues that have implications for how fluctuations in these proxies should be interpreted, and possibly weighted differently according some criteria. Before any such fluctuations are given any substantive interpretation, however, there are a host of methodological issues that have to be considered.

The most critical of these follow from the fact that the summation of the probabilities associated with the dates for the SPDs is not a statistically coherent procedure. This is because the summed probabilities can no longer been seen as probabilities, but rather represent the combination of events and uncertainties, making the two indistinguishable, and rendering the interpretation of the resulting sum difficult [@blackwell2003; @crema2022]. As @timpson2021[2] put it: 'the SPD is not *the* single best explanation of the data, nor even *a* explanation of the data, but rather a conflation of many possible explanations simultaneously, each of which is mired by the artefacts inherited from the calibration wiggles.' The SPD is not a model. It is the combined representation of a range of possible explanations for the data -- the frequency of dated events combined with the variable uncertainty associated with these [@carleton2021]. This means that an SPD cannot be directly analysed to draw inferences on population dynamics, nor can it be directly compared to other time-series data [@carleton2018; @timpson2021]. While this problem can never be entirely resolved, a range of approaches have been developed in an attempt to work around this issue.

The most commonly applied of these is a null-hypothesis significance testing approach by means of Monte Carlo simulation, as introduced by @shennan2013 and later expanded upon by @timpson2014. This works by comparing the observed RSPD with a series of simulated RSPDs, generated from a null-model. These null-models are typically a uniform, exponential or logistic distribution. These are chosen *a priori* with reference to common long-term population dynamics and are typically parametrised by fitting the models by means of regression or maximum likelihood estimation (MLE). The result from these simulations are then used to create a 95% critical envelope representing the null model. The proportion of the observed RSPD that falls outside this envelope is then used to estimate a global p-value indicating whether the null model can be rejected. In the case that it can, the portions of the observed RSPD that falls outside this envelope can subsequently be interpreted as representing potentially meaningful demographic events, relative to the null model. However, care has to be taken in how these are interpreted. First, this follows from the fact that 5% of the deviations from the critical envelope can be expected to be random, and there is no way to know which deviations this pertains to. Secondly, a for example exponential null model fit to the data is only one of an infinite set of exponential models with different growth rates that could be used. While a model fit by means of MLE will likely have a reasonable growth rate, and by extension exclude many other exponential fits as likely to explain the data, this can be difficult to determine. Finally, the p-value only indicates whether or not the null-model as a whole can be rejected as an explanation of the data, and does not provide statistical justification for interpreting local deviations themselves as meaningful demographic signals, as has often been the case [see review by @crema2022; @timpson2021].

@@

The procedure outlined above represent the standard approach when modelling population dynamics in archaeology. This can be helpful for determining whether the population behaviour has followed the trajectories of standard theoretical models, and is a useful first step here as few assumptions concerning the behaviour and overall development of number of shoreline dated sites over time could be made. However, standard models frequently used to evaluate RSPDs are consistently being rejected in the literature, and deviations from these rejected null-models do not in themselves allow for any interpretations of the deviations, nor a subsequent direct analysis of the SPDs themselves. As a result, and by improving on @goldstein2016, @timpson2021 have recently suggested an alternative framework for modelling past population dynamics by use of RSPDs, published with the accompanying R package  *ADMUR*. This involves fitting continuous piece-wise linear (CPL) models in addition to the standard models to the data, subjecting these to a procedure of model comparison, and finally evaluating the best model using a goodness-of-fit test in line with the Monte Carlo approach outlined above. The benefit of the CPL models is that these define discrete linear periods of development between hinge points that mark points in time in which the trajectory of the population proxy changes direction, thus allowing for the identification of key events with a conservative description of population dynamics between these. The approach thus also provides a clearer inferential justification for a further analysis and substantive interpretation of the identified best model. Following the evaluation of standard models as outlined above, the data were therefore subsequently subjected to the approach presented by @timpson2021.

@@

Sites surveyed by means of test-pitting between Horten and Arendal were initially retrieved from the national heritage database Askeladden [@directorate2018], totalling at `r nrow(surveyed)` records. The records were then manually reviewed and given a quality score based on the criteria in Table \@ref(tab:tab1), indicating the degree to which the spatial location and extent of the sites is believed to be represented in the geometries available in the database [see also @roalkvam2020]. All sites with a quality score of 4 or worse were excluded from further analysis. Any sites situated at elevations that result in a shoreline date earlier than 9469 BCE were then excluded. This marks the latest start-date among the employed displacement curves, and no sites are yet verifiably older than around 9300 BCE [@glorstad2016]. Data on excavated sites was originally compiled for @roalkvam2023 and has been compared with site data as listed in @damlien2021 and @nielsen2019. Only excavated sites with available spatial data in Askeladden or local databases at the Museum of Cultural History of the University of Oslo were included in the analysis. The 102 excavated sites in the dataset without relevant ^14^C-dates and that were originally shoreline dated in the reports were included in the SSPD along with the retained surveyed sites. This gave a total of  shoreline dated sites in the final SSPD.

The borders of the municipalities within which the shoreline dated sites are located were used to limit the radiocarbon sample. Radiocarbon dates were taken from @roalkvam2023 and @solheim. Dates done on food crusts were then excluded due the issue of marine reservoir effects [@nielsen2019, 83], resulting in a final total of ^14^C-dates. Following both from the point made by @freeman2018 concerning the comparison of radiocarbon dates from populations with different economic modes, as noted above, and from the fact that shoreline dating appears less reliable after 4000 BCE and appears to loose its utility c. 2500 BCE, the Mesolithic is the main focus of this analysis. Consequently, probabilities falling later than 2500 BCE were excluded from the analysis. Probabilities falling after 4000 BCE and the first, possible small-scale introduction of agriculture in south-eastern Norway were retained to account for edge-effects, but these results should be treated with care for the reasons noted above.

All analyses done in this study were performed using the R programming language [@rcoreteam]. Underlying data and programming code used for the paper is available in a version-controlled online repository at X. This is structured as a research compendium following @marwick2018b, to allow for reproducibility of the results [see also @Marwick2017]. Analysis of the ^14^C-dates were done using the R package *rcarbon* [@crema2021]. The R package *shoredate* was used for performing and handling the shoreline dating of sites [@roalkvam2023b]. Finally, the package *ADMUR* [@timpson2021] was used for model comparison.

@@

# Map prepared in 01_shoreline_spd.R 
knitr::include_graphics(here("analysis/figures/map1.png"))
```

@@

The RSPD was first subjected to the standard null-hypothesis testing approach through Monte Carlo simulation, as introduced above [see @shennan2013; @timpson2014], by fitting an exponential, logistic and uniform model to the observed RSPD. Following @timpson2021, the model parameters were identified by MLE search, using the differential evolution optimization algorithm *DEoptimR*. The Monte Carlo simulations were then performed using the *modelTest()* function from *rcarbon*. For each model, a series of individual calendar years are drawn from the model distribution, with replacement, the number of which equals the number of bins in the observed RSPD. These are then 'uncalibrated' to a single value on the ^14^C scale and a random error from among the observed errors is added to the date. These are then calibrated back to the calendar scale and finally summed. Here, this procedure was repeated 10000 times for each null model. The 2.5th and 97.5th quantile of the resulting summed probability for each year across all simulations are then retrieved to create the 95% critical envelope with which to compare the observed RSPD. The degree to which the observed RSPD deviates from the critical envelope is then used to calculate a global p-value, indicating whether or not the null can be rejected.

@@

Summing the probability of the shoreline dated sites and the model-fitting procedures followed the same structure as that for radiocarbon dates and was partly based on re-purposed programming code from *rcarbon*. However, idiosyncrasies in the dating method did necessitate some adjustments. To illustrate this, the procedure for shoreline dating a single site, as suggested in @roalkvam2023, is provided in Figure \@ref(fig:shoredate) and outlined below.

Four geological reconstructions of shoreline displacement in the region lays the foundation for the method as implemented here (Figure \@ref(fig:shoredate)B). These shoreline displacement curves are from Horten [@romundset2021], Porsgrunn [@sørensen2014; @sørensen2014b; @sorensen2023], Tvedestrand [@romundset2018; @romundset2018b] and Arendal [@romundset2018b], each associated with a shoreline isobase along which the trajectory for relative sea-level change has been the same [see @svendsen1987]. The first step in the dating procedure is to interpolate the shoreline displacement to the location to be dated. This is done by inverse distance weighting [e.g. @conolly2020], interpolating the relative sea-level change to the location of a site to be dated, based on it's distance from the isobases of the displacement curves.

```{r tpq_ranges, include = FALSE}
# Load data from mean_tpq_ranges to be used in paragraph below
load(here("analysis/data/derived_data/disp_tpq_ranges.RData"))
```

With an estimated trajectory of shoreline displacement at the site to be dated, the elevation of the site can be drawn on to find a likely date for when it was in use, under the assumption that it was in use when located on or close to the shoreline. Figure \@ref(fig:shoredate)C indicates where the elevation of the example site intersects the interpolated displacement curve. As it can be reasonably be assumed that a site was not in use when located under water, this effectively represents a *terminus post quem* (TPQ) date. However, a fairly common approach to shoreline dating has been find where the elevation of the site intersects the mean of the upper and lower limit of the displacement curve, and then adding a somewhat arbritrary error of $\pm$ 100 years to this point-estimate. However, finding TPQ dates that do not account for the likely distance between site and sea when the site was in use, limits, as with TPQ dates in general, the further inferential steps that can be taken. Furthermore, the more common practice of adding a constant error to a point-estimate does not take into account that the rate of shoreline displacement varies both spatially and temporally, and therefore introduces bias to the result on both the spatial and temporal scale, while also overestimating the precision of the method in most cases. This last point can be illustrated by finding the range between the upper and the lower limit of the four geological displacement curves for every meter from the marine limit and down to elevations that do not give an earliest date younger than 2500 BCE. This gives a mean TPQ range of `r round(tpqmean)` years with a standard deviation of `r round(tpqsd)` years across all four displacement curves, with this range varying significantly over time and between curves.

The analysis in @roalkvam2023 found that site phases ^14^C-dated to before 2500 BCE were found to have a likely elevation above sea-level that can be reasonably approximated by a gamma distribution with shape =  0.286 and scale = 20.833. This is incorporated in the procedure for shoreline dating used for this study (Figure \@ref(fig:shoredate)D). When performing this dating procedure, the gamma distribution is sequentially stepped through and transferred to the calendar scale by uniformly distributing the probability across the years in the range between the lower and upper limit of the interpolated displacement curve. This procedure thus gives the the shoreline date of a site when accounting for the likely elevation of the site above sea-level when it was in use [see @roalkvam2023 for details]. Given that the shoreline displacement curves have no inversions and should therefore before commutative [cf. @weninger2015, 545], each shoreline date is normalised to sum to unity. To reduce the computational cost of the simulation procedures to follow below, the gamma distribution is here stepped through at increments of 0.1 m and the calendar scale is kept at a resolution of 5 years.

```{r shoredate, echo = FALSE, fig.cap = "A) An example site (Askeladden ID 111967-1) relative to the isobases of the displacement curves.  B) The geological displacement curves in the region and the curve interpolated to the example site. C) The black line on the y-axis represents the elevation of the site. This crosses the interpolated displacement curve at 7680 and 7490 BCE, which thus in practice gives an estimate of when the location of the site emerged from the sea, giving a \\textit{terminus post quem} date. D) The shoreline dating approach used for this study, where the gamma distrbution on the y-axis describes the likely vertical relationship between the shoreline and the site when it was in use.",  out.width = "500px", fig.align="center", regfloat = TRUE}

knitr::include_graphics(here("analysis/figures/example_shoredate.png"))
```

The implications of different approaches to shoreline dating and summing the results is further illustrated in Figure \@ref(fig:sumshore). In Figure \@ref(fig:sumshore)A  the sites to be shoreline dated are plotted according to their elevation. Due to the variable rates of shoreline displacement within the study region, the same elevation does not necessarily equate to the same date, and so this does not directly tell us much about their temporal distribution. For example, as the trajectory for sea-level regression is starker towards the north-east, the same elevation at a location to the south-west implies a younger date that one towards the north-east.

When the frequency distribution of shoreline dates have been investigated in the past, this has typically involved finding point-estimates of dates by either using the mean of displacement curves, or by taking mean of the TPQ dates for each site (as illustrated in Figure \@ref(fig:shoredate)C). These point-estimates are then aggregated in bins of 200 or 500 years [e.g. @breivik2014; @solheim2018; @fossum2020; @mjærum2022; @tallavaara2020]. This approach is illustrated in Figure \@ref(fig:sumshore)B. The SSPD resulting from instead using the probabilistic method of shoreline dating, given in Figure \@ref(fig:sumshore)C, clearly demonstrates that accounting for the site-sea relationship gives a substantially different result compared to the more traditional approach, which is clearly indicated by the shifts in the positions of peaks and troughs in the distribution after c. 8000 BCE.

```{r sumshore, echo = FALSE, fig.cap = "A) The distribution of the shoreline dated sites (n = 921) by elevation. The binwidth is 4 m. B) The mean of \\textit{terminus post quem} dates for the sites, plotted with a binsize of 200 years, reflecting the common approach to summing shoreline dates. C) The result of instead summing the probability of the shoreline dates achievied with the method employed here, which accounts for the likely distance between sites and shoreline. The binwidth is 5 years, following the resolution used with the dating procedure.",  out.width = "500px", fig.align="center", regfloat = TRUE}

knitr::include_graphics(here("analysis/figures/summed_examples.png"))
```

@@

```{r incpolys, echo = FALSE, fig.cap = "A) Density of included surveyed sites and excavated sites dated by means of shoreline dating (n = 921) as distributed across 2 km wide line segments that run perpendicular to the shoreline gradient. B) The displacement curves interpolated to the centre of each segment for use in the Monte Carlo simulations below.", out.width = "500px", fig.align="center", regfloat = TRUE}
knitr::include_graphics(here("analysis/figures/incpolys.png"))
```

Following from the commutative nature of the shoreline dates dealt with here, the Monte Carlo simulation with the SSPD is here based on drawing a sample of calendar dates from the observed date range in the SSPD, equalling the number of shoreline dated sites, where the probability of drawing any individual year is determined by the null model of choice. This is equivalent to the sampling method *calsample* from *rcarbon* [@crema2021]. Each sampled date is then assigned one of the pre-interpolated displacement curves, the probability of which is weighted by the density of observed sites within each 2 km interval. The calendar date is then 'uncalibrated' -- to follow the RSPD terminology -- to an elevation range from which a single elevation value is then drawn with uniform probability between the lower and upper limit of the displacement curve at that calendar year, using intervals of 5 cm. As shoreline dating is here done with a gamma distribution with the same parameters across all sites, there is no equivalent of the error term for ^14^C-dates that determine the shape of the Gaussian distribution associated with the ^14^C age. Consequently, the elevation value retrieved by 'uncalibrating' the shoreline date was then shoreline dated using the displacement curve for the relevant 2 km interval with the same gamma distribution for all samples. Having dated the number of samples equalling the number of shoreline dated sites, these were then summed, and the entire process repeated a total of 10000 times. The 97.5% highest and 2.5% lowest summed probability at each interval of 5 years across all simulations was then retrieved to create the 95% critical envelope, and the global p-value found in the same manner as for the RSPDs. This was done using re-purposed code from *rcarbon*.

## Continous piece-wise linear models
The procedures outlined above represent standard approaches when modelling population dynamics in archaeology. These can be helpful for determining whether the population behaviour has followed the trajectories of common theoretical population models, and is an especially useful exercise here as few assumptions concerning the behaviour and overall development of number of shorebound sites over time could be made. However, standard models frequently used to evaluate RSPDs are consistently being rejected in the literature, and deviations from these rejected null-models do not in themselves allow for any interpretations of the deviations, nor a subsequent direct analysis of the SPDs -- although this has frequently been the case. As a result, @timpson2021 have recently suggested an alternative framework for modelling past population dynamics. This involves fitting continuous piece-wise linear (CPL) models in addition to the standard models to the data, subjecting these to a procedure for model comparison, and finally testing the best model using the approach outlined above. The benefit of the CPL models is that these define discrete periods between hinge points that marks points in time in which the trajectory of relative population size changes direction, thus identifying key events and characteristics of the population dynamics between these. Following the evaluation of standard models outlined above, the data was therefore subsequently subjected to the approach outlined by @timpson2021.
